---
title: "classification_tree"
author: "Weihan Liu"
date: "06/12/2019"
output: html_document
---

categories of trees
1.Classification vs regression trees
2.bagged trees
3.random forest
4.boosted trees(GBM)

trees requires NO normalization or standarization for numeric columns, can handle both numeric and categorical columns at the same time
trees require very little data preprocessing(almost none). However, trees tend to have high variace and tend to overfit(too specific), so you need to be careful to tune the parameters/hyperparameters sufficiently


##classification tree

Let's get started and build our first classification tree. A classification tree is a decision tree that performs a classification (vs regression) task.

You will train a decision tree model to understand which loan applications are at higher risk of default using a subset of the German Credit Dataset. The response variable, called "default", indicates whether the loan went into a default or not, which means this is a binary classification problem (there are just two classes).
link to the dataset we use: https://archive.ics.uci.edu/ml/datasets/Statlog+%28German+Credit+Data%29

### fit the decision tree and the rpart.plot package to visualize the tree.
```{r}
# Look at the data
str(creditsub)

# Create the model, provide the training data frame
credit_model <- rpart(formula = default ~ ., 
                      data = creditsub, 
                      method = "class")

# Display the results
rpart.plot(x = credit_model, yesno = 2, type = 0, extra = 0)
```



###train/test split
For this exercise, you'll randomly split the German Credit Dataset into two pieces: a training set (80%) called credit_train and a test set (20%) that we will call credit_test. We'll use these two sets throughout the chapter.
```{r}
# Total number of rows in the credit data frame
n <- nrow(credit)

# Number of rows for the training set (80% of the dataset)
n_train <- round(0.8 * n) 

# Create a vector of indices which is an 80% random sample. sample(data,size)
set.seed(123)
train_indices <- sample(1:n, n_train)

# Subset the credit data frame to training indices only
credit_train <- credit[train_indices, ]  
  
# Exclude the training indices to create the test set
credit_test <- credit[-train_indices, ]  
```


###Train a classification tree model
In this exercise, you will train a model on the newly created training set and print the model object to get a sense of the results.
```{r}
# Train the model (to predict 'default')
credit_model <- rpart(formula = default ~ ., 
                      data = credit_train, 
                      method = "class")

# Look at the model output                      
print(credit_model)
```


There are several metrics to evaluate the performance of classification models
1.Accuracy
2.Confusion Matrix
3.Log-Loss
4.AUC

###compute a confusion matrix
Compute confusion matrix
As discussed in the previous video, there are a number of different metrics by which you can measure the performance of a classification model. In this exercise, we will evaluate the performance of the model using test set classification error. A confusion matrix is a convenient way to examine the per-class error rates for all classes at once.

The confusionMatrix() function from the caret package prints both the confusion matrix and a number of other useful classification metrics such as "Accuracy" (fraction of correctly classified instances).

The caret package contains the confusionMatrix function
```{r}
library(caret)
# Generate predicted classes using the model object
class_prediction <- predict(object = credit_model,  
                        newdata = credit_test,   
                        type = "class")  
                            
# Calculate the confusion matrix for the test set
confusionMatrix(data = class_prediction,       
                reference = credit_test$default)  
```


Another aspect of performace measufe is that we want the resulting leafs to be as pure/homogeneous as possible. You can use gini index to measur eth puriy of the leafs

###Compare models with a different splitting criterion, gini vs entropy
Train two models that use a different splitting criterion and use the validation set to choose a "best" model from this group. To do this you'll use the parms argument of the rpart() function. This argument takes a named list that contains values of different parameters you can use to change how the model is trained. Set the parameter split to control the splitting criterion.

using gini index vs information gain(entropy) as the splitting cretaria: http://www.learnbymarketing.com/481/decision-tree-flavors-gini-info-gain/
```{r}
# Train a gini-based model
credit_model1 <- rpart(formula = default ~ ., 
                       data = credit_train, 
                       method = "class",
                       parms = list(split = "gini"))

# Train an information-based model
credit_model2 <- rpart(formula = default ~ ., 
                       data = credit_train, 
                       method = "class",
                       parms = list(split = "information"))

# Generate predictions on the validation set using the gini model
pred1 <- predict(object = credit_model1, 
             newdata = credit_test,
             type = "class")    

# Generate predictions on the validation set using the information model
pred2 <- predict(object = credit_model2, 
             newdata = credit_test,
             type = "class")

# Compare classification error
ce(actual = credit_test$default, 
   predicted = pred1)
ce(actual = credit_test$default, 
   predicted = pred2)  
```



##Regression trees
We use regression tree to predict continuous value of a variable, we use classification tree to predict categorical value of a variable

Types of data
1.Training set
2.Validation set: tune hyperparameters and choose the best candidate model
3.Test set: meant only to be used once, at the end of your pipeline, cannot be used to tune hyperparameters

###Split the data
These examples will use a subset of the Student Performance Dataset from UCI ML Dataset Repository:https://archive.ics.uci.edu/ml/datasets/Student+Performance
The goal of this exercise is to predict a student's final Mathematics grade based on the following variables: sex, age, address, studytime (weekly study time), schoolsup (extra educational support), famsup (family educational support), paid (extra paid classes within the course subject) and absences.
The response is final_grade (numeric: from 0 to 20, output target).
After initial exploration, split the data into training, validation, and test sets. In this chapter, we will introduce the idea of a validation set, which can be used to select a "best" model from a set of competing models.
In Chapter 1, we demonstrated a simple way to split the data into two pieces using the sample() function. In this exercise, we will take a slightly different approach to splitting the data that allows us to split the data into more than two parts (here, we want three: train, validation, test). We still use the sample() function, but instead of sampling the indices themselves, we will assign each row to either the training, validation or test sets according to a probability distribution.
The dataset grade is already in your workspace.

```{r}
# Look at the data
str(grade)
 
# Set seed and create assignment
set.seed(1)
assignment <- sample(1:3, size = nrow(grade), prob = c(0.7,0.15,0.15), replace = TRUE)

# Create a train, validation and tests from the original data frame 
grade_train <- grade[assignment == 1, ]    # subset grade to training indices only
grade_valid <- grade[assignment == 2, ]  # subset grade to validation indices only
grade_test <- grade[assignment == 3, ]   # subset grade to test indices only
```


###Train a regression tree 
In this exercise, we will use the grade_train dataset to fit a regression tree using rpart() and visualize it using rpart.plot(). A regression tree plot looks identical to a classification tree plot, with the exception that there will be numeric values in the leaf nodes instead of predicted classes.

This is very similar to what we did previously in Chapter 1. When fitting a classification tree, we use method = "class", however, when fitting a regression tree, we need to set method = "anova". By default, the rpart() function will make an intelligent guess as to what the method value should be based on the data type of your response column, but it's recommened that you explictly set the method for reproducibility reasons (since the auto-guesser may change in the future).

The grade_train training set is loaded into the workspace.
```{r}
# Train the model
grade_model <- rpart(formula = final_grade ~ ., 
                     data = grade_train, 
                     method = "anova")

# Look at the model output                      
print(grade_model)

# Plot the tree model
rpart.plot(x = grade_model, yesno = 2, type = 0, extra = 0)
```


###evaluate the performance of a regression tree
Predict the final grade for all students in the test set. The grade is on a 0-20 scale. Evaluate the model based on test set RMSE (Root Mean Squared Error). RMSE tells us approximately how far away our predictions are from the true values.
```{r}
# Generate predictions on a test set
pred <- predict(object = grade_model,   # model object 
                newdata =grade_test)  # test dataset

# Compute the RMSE
rmse(actual = grade_test$final_grade, 
     predicted = pred)
```


###hyperparameters for a decision tree
hyperparameters cannot be learned by fitting the models. You shoud set hyperparameters before training the model. Important hyperparameters for a tree:
1.Minsplit: minimum number of data points required to attempt a split
2.cp: complexity parameter: a penalty term for tree size. the smaller the cp, the more complex the tree will be
3.maxdepth: depth of a decision tree, maximum number of nodes between a leaf node and root node

Tuning the model
Tune (or "trim") the model using the prune() function by finding the best "CP" value (CP stands for "Complexity Parameter").
```{r}
# Plot the "CP Table"
plotcp(grade_model)

# Print the "CP Table"
print(grade_model$cptable)

# Retrieve optimal cp value based on cross-validated error
opt_index <- which.min(grade_model$cptable[, "xerror"])
cp_opt <- grade_model$cptable[opt_index, "CP"]

# Prune the model (to optimized cp value)
grade_model_opt <- prune(tree = grade_model, cp = cp_opt)
                          
# Plot the optimized model
rpart.plot(x = grade_model_opt, yesno = 2, type = 0, extra = 0)
```


Grid search: a brute force search of all combination of model hyperparameters(each combination will give a slightly different model), to return the best combination of model hyperparameters.
You need to choose a performance matrics to evaluate the best-performing model

Generate a grid of hyperparameter values
Use expand.grid() to generate a grid of maxdepth and minsplit values.
```{r}
# Establish a list of possible values for minsplit and maxdepth
minsplit <- seq(1, 4, 1)
maxdepth <- seq(1, 6, 1)

# Create a data frame containing all combinations 
hyper_grid <- expand.grid(minsplit = minsplit, maxdepth = maxdepth)

# Check out the grid
head(hyper_grid)

# Print the number of grid combinations
nrow(hyper_grid)
```


Generate a grid of models
In this exercise, we will write a simple loop to train a "grid" of models and store the models in a list called grade_models. R users who are familiar with the apply functions in R could think about how this loop could be easily converted into a function applied to a list as an extra-credit thought experiment.
```{r}
# Number of potential models in the grid
num_models <- nrow(hyper_grid)

# Create an empty list to store models
grade_models <- list()

# Write a loop over the rows of hyper_grid to train the grid of models
for (i in 1:num_models) {

    # Get minsplit, maxdepth values at row i
    minsplit <- hyper_grid$minsplit[i]
    maxdepth <- hyper_grid$maxdepth[i]

    # Train a model and store in the list
    grade_models[[i]] <- rpart(formula = final_grade ~ ., 
                               data = grade_train, 
                               method = "anova",
                               minsplit = minsplit,
                               maxdepth = maxdepth)
}
```

Evaluate the grid
Earlier in the chapter we split the dataset into three parts: training, validation and test.

A dataset that is not used in training is sometimes referred to as a "holdout" set. A holdout set is used to estimate model performance and although both validation and test sets are considered to be holdout data, there is a key difference:

Just like a test set, a validation set is used to evaluate the performance of a model. The difference is that a validation set is specifically used to compare the performance of a group of models with the goal of choosing a "best model" from the group. All the models in a group are evaluated on the same validation set and the model with the best performance is considered to be the winner.
Once you have the best model, a final estimate of performance is computed on the test set.
A test set should only ever be used to estimate model performance and should not be used in model selection. Typically if you use a test set more than once, you are probably doing something wrong.
```{r}
# Number of potential models in the grid
num_models <- length(grade_models)

# Create an empty vector to store RMSE values
rmse_values <- c()

# Write a loop over the models to compute validation RMSE
for (i in 1:num_models) {

    # Retrieve the i^th model from the list
    model <- grade_models[[i]]
    
    # Generate predictions on grade_valid 
    pred <- predict(object = model,
                    newdata = grade_valid)
    
    # Compute validation RMSE and add to the 
    rmse_values[i] <- rmse(actual = grade_valid$final_grade, 
                           predicted = pred)
}

# Identify the model with smallest validation set RMSE
best_model <- grade_models[[which.min(rmse_values)]]

# Print the model paramters of the best model. All model parameters are stored in "control"
best_model$control

# Compute test set RMSE on best_model
pred <- predict(object = best_model,
                newdata = grade_test)
rmse(actual = grade_test$final_grade, 
     predicted = pred)
```




##bagged trees

bagged trees combine multiple trees to reduce variance and increase accuracy. Big variance is a main issue for deciision tree

Train a bagged tree model
Let's start by training a bagged tree model. You'll be using the bagging() function from the ipred package. The number of bagged trees can be specified using the nbagg parameter, but here we will use the default (25).

If we want to estimate the model's accuracy using the "out-of-bag" (OOB) samples, we can set the the coob parameter to TRUE. The OOB samples are the training obsevations that were not selected into the bootstrapped sample (used in training). Since these observations were not used in training, we can use them instead to evaluate the accuracy of the model (done automatically inside the bagging() function)

Here we will used the credit dataset, note this is a classification tree, not a regression tree
```{r}
# Bagging is a randomized model, so let's set a seed (123) for reproducibility
set.seed(123)

# Train a bagged model. Note the bagging here works exactly the same as predict() function we previously used for single tree
credit_model <- bagging(formula = default ~ ., 
                        data = credit_train,
                        coob = TRUE)

# Print the model
print(credit_model)
```


Evaluate the performance of bagged tree models
ROC curve is used to compare the performance across different classifiers. AUC is used to quantify the performance of a model. AUC of 0.5 means the classifier is no better than random chance. AUC = 0 means it classifies everything wrong, AUC = 1 means it classify everything correctly.
```{r}
# Generate predicted classes using the model object, here is a single classification tree instead of a bagged tree.
class_prediction <- predict(object = credit_model,    
                            newdata = credit_test,  
                            type = "class")  # return classification labels

# Print the predicted classes
print(class_prediction)

# Calculate the confusion matrix for the test set
confusionMatrix(data = class_prediction,       
                reference = credit_test$default)  
```



Prediction and confusion matrix
As you saw in the video, a confusion matrix is a very useful tool for examining all possible outcomes of your predictions (true positive, true negative, false positive, false negative).

In this exercise, you will predict those who will default using bagged trees. You will also create the confusion matrix using the confusionMatrix() function from the caret package.

It's always good to take a look at the output using the print() function.
```{r}
# Generate predictions on the test set
pred <- predict(object = credit_model,
                newdata = credit_test,
                type = "prob")

# `pred` is a matrix
class(pred)
                
# Look at the pred format
head(pred)
                
# Compute the AUC (`actual` must be a binary (or 1/0 numeric) vector)
auc(actual = ifelse(credit_test$default == "yes", 1, 0), 
    predicted = pred[,"yes"])                    
```
We usually use AUC or confusion martrix to evaluate the perforance of classification, RMSE to measure regression questions


Cross validation: reduce variance(we will use Caret package here)
K-fold cross validation: partition the rows into k fold subset of equal sizes. 1 subset will be used as test set, the rest are used as training set
Cross-validate a bagged tree model in caret
Use caret::train() with the "treebag" method to train a model and evaluate the model using cross-validated AUC. The caret package allows the user to easily cross-validate any model across any relevant performance metric. In this case, we will use 5-fold cross validation and evaluate cross-validated AUC (Area Under the ROC Curve).
```{r}
# Specify the training configuration
ctrl <- trainControl(method = "cv",     # Cross-validation
                     number = 5,      # 5 folds
                     classProbs = TRUE,                  # For AUC
                     summaryFunction = twoClassSummary)  # For AUC

# Cross validate the credit model using "treebag" method; 
# Track AUC (Area under the ROC curve)
set.seed(1)  # for reproducibility
credit_caret_model <- train(default ~ .,   
                            data = credit_train, 
                            method = "treebag",
                            metric = "ROC",
                            trControl = ctrl)#train() function here is equivalent to rpart() function we previously used in rpart package

# Look at the model object
print(credit_caret_model)

# Inspect the contents of the model list 
names(credit_caret_model)

# Print the CV AUC
credit_caret_model$results[,"ROC"]
```


Generate predictions from the caret model
Generate predictions on a test set for the caret model.
```{r}
# Generate predictions on the test set
pred <- predict(object = credit_caret_model, 
                newdata = credit_test,
                type = "prob")

# Compute the AUC (`actual` must be a binary (or 1/0 numeric) vector)
auc(actual = ifelse(credit_test$default == "yes", 1, 0), 
                    predicted = pred[,"yes"])
```


Compare test set performance to CV performance
In this excercise, you will print test set AUC estimates that you computed in previous exercises. These two methods use the same code underneath, so the estimates should be very similar.

The credit_ipred_model_test_auc object stores the test set AUC from the model trained using the ipred::bagging() function.
The credit_caret_model_test_auc object stores the test set AUC from the model trained using the caret::train() function with method = "treebag".
Lastly, we will print the 5-fold cross-validated estimate of AUC that is stored within the credit_caret_model object. This number will be a more accurate estimate of the true model performance since we have averaged the performance over five models instead of just one.

On small datasets like this one, the difference between test set model performance estimates and cross-validated model performance estimates will tend to be more pronounced. When using small data, it's recommended to use cross-validated estimates of performance because they are more stable.
```{r}
# Print ipred::bagging test set AUC estimate
print(credit_ipred_model_test_auc)

# Print caret "treebag" test set AUC estimate
print(credit_caret_model_test_auc)
                
# Compare to caret 5-fold cross-validated AUC
credit_caret_model$results[, "ROC"]
```





###Random Forest: an improvedment on bagged trees
For random forest, at each split, only a subset of variables are randomly selected. While for previously mentioned tree algorithms, they use all the variables. This makes sure the ensemble doesn't reply too heavily on a single prefictable variable, and make fair use of all the predictive variables.

Train a Random Forest model
Here you will use the randomForest() function from the randomForest package to train a Random Forest classifier to predict loan default.
```{r}
# Train a Random Forest
set.seed(1)  # for reproducibility
credit_model <- randomForest(formula = default ~ ., 
                             data = credit_train)
                             
# Print the model output                             
print(credit_model)
```


The samples that are not selected by the bootstrap in random forest is called OOB
OOB is the mean prediction error on each training sample xᵢ, using only the trees that did not have xᵢ in their bootstrap sample.

Evaluate out-of-bag error
Here you will plot the OOB error as a function of the number of trees trained, and extract the final OOB error of the Random Forest model from the trained model object.
```{r}
# Grab OOB error matrix & take a look
err <- credit_model$err.rate
head(err)

# Look at final OOB error rate (last row in err matrix)
oob_err <- err[nrow(err), "OOB"]
print(oob_err)

# Plot the model trained in the previous exercise
plot(credit_model)

# Add a legend since it doesn't have one by default
legend(x = "right", 
       legend = colnames(err),
       fill = 1:ncol(err))
```


Evaluate model performance on a test set
Use the caret::confusionMatrix() function to compute test set accuracy and generate a confusion matrix. Compare the test set accuracy to the OOB accuracy.
```{r}
# Generate predicted classes using the model object
class_prediction <- predict(object = credit_model,   # model object 
                            newdata = credit_test,  # test dataset
                            type = "class") # return classification labels
                            
# Calculate the confusion matrix for the test set
cm <- confusionMatrix(data = class_prediction,       # predicted classes
                      reference = credit_test$default)  # actual classes
print(cm)

# Compare test set accuracy to OOB accuracy
paste0("Test Accuracy: ", cm$overall[1])
paste0("OOB Accuracy: ", 1 - oob_err)
```


Evaluate test set AUC
In Chapter 3, we learned about the AUC metric for evaluating binary classification models. In this exercise, you will compute test set AUC for the Random Forest model.
```{r}
# Generate predictions on the test set
pred <- predict(object = credit_model,
            newdata = credit_test,
            type = "prob")

# `pred` is a matrix
class(pred)
                
# Look at the pred format
head(pred)
                
# Compute the AUC (`actual` must be a binary 1/0 numeric vector)
auc(actual = ifelse(credit_test$default == "yes", 1, 0), 
    predicted = pred[,"yes"])                    
```



Tuning a random forest model
important random forest hyperparameters:
1.ntree: number of trees, default 500

the following two parameters controls randomness of a random forest model
2.mtry: number of variables randomly sampled as candidate at each split
3.samplesize:number of samples to train on(default is 62.3% because this is the proportion of samples expected to be unique in a bootstrp sample)

The following two parameters control the complexity of tree
4.nodesize:minimum size of the terminal nodes
5.maxnodes: maximum number of terminal nodes


mtry is the most important hyperparameter, and we can use tuneRF() function to tune this hyperparameter

Tuning a Random Forest via mtry
In this exercise, you will use the randomForest::tuneRF() to tune mtry (by training several models). This function is a specific utility to tune the mtry parameter based on OOB error, which is helpful when you want a quick & easy way to tune your model. A more generic way of tuning Random Forest parameters will be presented in the following exercise.
```{r}
# Use the tuneRF() function in place of the randomForest() function to train a series of models with different mtry values and examine the the results.
# Note that (unfortunately) the tuneRF() interface does not support the typical formula input that we've been using, but instead uses two arguments, x (matrix or data frame of predictor variables) and y (response vector; must be a factor for classification).
# Execute the tuning process
set.seed(1)              
res <- tuneRF(x = subset(credit_train, select = -default),
              y = credit_train$default,
              ntreeTry = 500)
               
# Look at results
print(res)

# Find the mtry value that minimizes OOB Error
mtry_opt <- res[,"mtry"][which.min(res[,"OOBError"])]
print(mtry_opt)

# If you just want to return the best RF model (rather than results)
# you can set `doBest = TRUE` in `tuneRF()` to return the best RF model
# instead of a set performance matrix.
```


Tuning a Random Forest via tree depth
In Chapter 2, we created a manual grid of hyperparameters using the expand.grid() function and wrote code that trained and evaluated the models of the grid in a loop. In this exercise, you will create a grid of mtry, nodesize and sampsize values. In this example, we will identify the "best model" based on OOB error. The best model is defined as the model from our grid which minimizes OOB error.

Keep in mind that there are other ways to select a best model from a grid, such as choosing the best model based on validation AUC. However, for this exercise, we will use the built-in OOB error calculations instead of using a separate validation set.

```{r}
# Establish a list of possible values for mtry, nodesize and sampsize
mtry <- seq(4, ncol(credit_train) * 0.8, 2)
nodesize <- seq(3, 8, 2)
sampsize <- nrow(credit_train) * c(0.7, 0.8)

# Create a data frame containing all combinations 
hyper_grid <- expand.grid(mtry = mtry, nodesize = nodesize, sampsize = sampsize)

# Create an empty vector to store OOB error values
oob_err <- c()

# Write a loop over the rows of hyper_grid to train the grid of models
for (i in 1:nrow(hyper_grid)) {

    # Train a Random Forest model
    model <- randomForest(formula = default ~ ., 
                          data = credit_train,
                          mtry = hyper_grid$mtry[i],
                          nodesize = hyper_grid$nodesize[i],
                          sampsize = hyper_grid$sampsize[i])
                          
    # Store OOB error for the model                      
    oob_err[i] <- model$err.rate[nrow(model$err.rate), "OOB"]
}

# Identify optimal set of hyperparmeters based on OOB error
opt_i <- which.min(oob_err)
print(hyper_grid[opt_i,])
```


Boosting:

diff between boosting and bagging: boosted tree learn from previous trees, bagged tree are just dividing features and learn.
explanation of bagging vs boosting: https://towardsdatascience.com/basic-ensemble-learning-random-forest-adaboost-gradient-boosting-step-by-step-explained-95d49d1e2725
two popular boosting algorithms:
1.Adaboost
2.GBM

1.Adaboost algorithm:
a.Train decision tree with equally weighted observations
b.Increase/Lower the weights of observations. Observations that are easy to classify get lower weight, observations that are hard to classify get higher weight
c.second tree is grown on weighted data
d.New model: Tree 1 + Tree 2
e.Classification error from this new 2-tree emsemble model
f.grow 3rd tree to predict the revised residuals
g.repeat this process for a specified number of iteratioms


Train a GBM model
Here you will use the gbm() function to train a GBM classifier to predict loan default. You will train a 10,000-tree GBM on the credit_train dataset, which is pre-loaded into your workspace.

Using such a large number of trees (10,000) is probably not optimal for a GBM model, but we will build more trees than we need and then select the optimal number of trees based on early performance-based stopping. The best GBM model will likely contain fewer trees than we started with.

For binary classification, gbm() requires the response to be encoded as 0/1 (numeric), so we will have to convert from a "no/yes" factor to a 0/1 numeric response column.

Also, the the gbm() function requires the user to specify a distribution argument. For a binary classification problem, you should set distribution = "bernoulli". The Bernoulli distribution models a binary response.
```{r}
# Convert "yes" to 1, "no" to 0
credit_train$default <- ifelse(credit_train$default == "yes", 1, 0)

# Train a 10000-tree GBM model
set.seed(1)
credit_model <- gbm(formula = default ~ ., 
                    distribution = "bernoulli", 
                    data = credit_train,
                    n.trees = 10000)
                    
# Print the model object                    
print(credit_model)

# summary() prints variable importance
summary(credit_model)
```



Prediction using a GBM model
The gbm package uses a predict() function to generate predictions from a model, similar to many other machine learning packages in R. When you see a function like predict() that works on many different types of input (a GBM model, a RF model, a GLM model, etc), that indicates that predict() is an "alias" for a GBM-specific version of that function. The GBM specific version of that function is predict.gbm(), but for convenience sake, we can just use predict() (either works).

One thing that's particular to the predict.gbm() however, is that you need to specify the number of trees used in the prediction. There is no default, so you have to specify this manually. For now, we can use the same number of trees that we specified when training the model, which is 10,000 (though this may not be the optimal number to use).

Another argument that you can specify is type, which is only relevant to Bernoulli and Poisson distributed outcomes. When using Bernoulli loss, the returned value is on the log odds scale by default and for Poisson, it's on the log scale. If instead you specify type = "response", then gbm converts the predicted values back to the same scale as the outcome. This will convert the predicted values into probabilities for Bernoulli and expected counts for Poisson.
```{r}
# Since we converted the training response col, let's also convert the test response col
credit_test$default <- ifelse(credit_test$default == "yes", 1, 0)

# Generate predictions on the test set
preds1 <- predict(object = credit_model, 
                  newdata = credit_test,
                  n.trees = 10000)

# Generate predictions on the test set (scale to response)
preds2 <- predict(object = credit_model, 
                  newdata = credit_test,
                  n.trees = 10000,
                  type = "response")

# Compare the range of the two sets of predictions
range(preds1)
range(preds2)
```


Evaluate test set AUC
Compute test set AUC of the GBM model for the two sets of predictions. We will notice that they are the same value. That's because AUC is a rank-based metric, so changing the actual values does not change the value of the AUC.

However, if we were to use a scale-aware metric like RMSE to evaluate performance, we would want to make sure we converted the predictions back to the original scale of the response.
```{r}
# Generate the test set AUCs using the two sets of preditions & compare
auc(actual = credit_test$default, predicted = preds1)  #default
auc(actual = credit_test$default, predicted = preds2)  #rescaled
```



GBM hyperparameters
hyperparameter tuning is especially important for GBM

1)n.tree: number of trees
2)bag.fraction: proportion of observations to be sampled in each tree
3)n.minobsinnode: minimum number of observations in the tree terminal nodes
4).interaction depth: maximum nodes per tree
5).shrinkage: learning rate


Early stopping in GBMs
Use the gbm.perf() function to estimate the optimal number of boosting iterations (aka n.trees) for a GBM model object using both OOB and CV error. When you set out to train a large number of trees in a GBM (such as 10,000) and you use a validation method to determine an earlier (smaller) number of trees, then that's called "early stopping". The term "early stopping" is not unique to GBMs, but can describe auto-tuning the number of iterations in an iterative learning algorithm.
```{r}
# Optimal ntree estimate based on OOB
ntree_opt_oob <- gbm.perf(object = credit_model, 
                          method = "OOB", 
                          oobag.curve = TRUE)

# Train a CV GBM model
set.seed(1)
credit_model_cv <- gbm(formula = default ~ ., 
                       distribution = "bernoulli", 
                       data = credit_train,
                       n.trees = 10000,
                       cv.folds = 2)

# Optimal ntree estimate based on CV
ntree_opt_cv <- gbm.perf(object = credit_model_cv, 
                         method = "cv")
 
# Compare the estimates                         
print(paste0("Optimal n.trees (OOB Estimate): ", ntree_opt_oob))                         
print(paste0("Optimal n.trees (CV Estimate): ", ntree_opt_cv))
```


OOB vs CV-based early stopping
In the previous exercise, we used OOB error and cross-validated error to estimate the optimal number of trees in the GBM. These are two different ways to estimate the optimal number of trees, so in this exercise we will compare the performance of the models on a test set. We can use the same model object to make both of these estimates since the predict.gbm() function allows you to use any subset of the total number of trees (in our case, the total number is 10,000).

Instructions
```{r}
# Generate predictions on the test set using ntree_opt_oob number of trees
preds1 <- predict(object = credit_model, 
                  newdata = credit_test,
                  n.trees = ntree_opt_oob)
                  
# Generate predictions on the test set using ntree_opt_cv number of trees
preds2 <- predict(object = credit_model, 
                  newdata = credit_test,
                  n.trees = ntree_opt_cv)   

# Generate the test set AUCs using the two sets of preditions & compare
auc1 <- auc(actual = credit_test$default, predicted = preds1)  #OOB
auc2 <- auc(actual = credit_test$default, predicted = preds2)  #CV 

# Compare AUC 
print(paste0("Test set AUC (OOB): ", auc1))                         
print(paste0("Test set AUC (CV): ", auc2))
```


Using ROC curve to evaluate performance of all models 
Compare all models based on AUC
In this final exercise, we will perform a model comparison across all types of models that we've learned about so far: Decision Trees, Bagged Trees, Random Forest and Gradient Boosting Machine (GBM). The models were all trained on the same training set, credit_train, and predictions were made for the credit_test dataset.

We have pre-loaded four sets of test set predictions, generated using the models we trained in previous chapters (one for each model type). The numbers stored in the prediction vectors are the raw predicted values themselves -- not the predicted class labels. Using the raw predicted values, we can calculate test set AUC for each model and compare the results.
```{r}
# Generate the test set AUCs using the two sets of predictions & compare
actual <- credit_test$default
dt_auc <- auc(actual = actual, predicted = dt_preds) #simple decision tree
bag_auc <- auc(actual = actual, predicted = bag_preds) #bagged tree
rf_auc <- auc(actual = actual, predicted = rf_preds) #random forest
gbm_auc <- auc(actual = actual, predicted = gbm_preds)#gbm 

# Print results
sprintf("Decision Tree Test AUC: %.3f", dt_auc)
sprintf("Bagged Trees Test AUC: %.3f", bag_auc)
sprintf("Random Forest Test AUC: %.3f", rf_auc)
sprintf("GBM Test AUC: %.3f", gbm_auc)
```


Plot & compare ROC curves
We conclude this course by plotting the ROC curves for all the models (one from each chapter) on the same graph. The ROCR package provides the prediction() and performance() functions which generate the data required for plotting the ROC curve, given a set of predictions and actual (true) values.

The more "up and to the left" the ROC curve of a model is, the better the model. The AUC performance metric is literally the "Area Under the ROC Curve", so the greater the area under this curve, the higher the AUC, and the better-performing the model is.

The ROCR package can plot multiple ROC curves on the same plot if you plot several sets of predictions as a list.

The prediction() function takes as input a list of prediction vectors (one per model) and a corresponding list of true values (one per model, though in our case the models were all evaluated on the same test set so they all have the same set of true values). The prediction() function returns a "prediction" object which is then passed to the performance() function.
The performance() function generates the data necessary to plot the curve from the "prediction" object. For the ROC curve, you will also pass along two measures, "tpr" and "fpr".
Once you have the "performance" object, you can plot the ROC curves using the plot() method. We will add some color to the curves and a legend so we can tell which curves belong to which algorithm.
```{r}
# List of predictions
preds_list <- list(dt_preds, bag_preds, rf_preds, gbm_preds)

# List of actual values (same for all)
m <- length(preds_list)
actuals_list <- rep(list(credit_test$default), m)

# Plot the ROC curves
pred <- prediction(preds_list, actuals_list)
rocs <- performance(pred, "tpr", "fpr")
plot(rocs, col = as.list(1:m), main = "Test Set ROC Curves")
legend(x = "bottomright", 
       legend = c("Decision Tree", "Bagged Trees", "Random Forest", "GBM"),
       fill = 1:m)
```

