---
title: "Chr7 ML TS Gene Gviz Fig"
author: "Shirley Zhou"
date: "5/28/2020"
output: html_document
---

Load the following package needed:
1) Gviz to perform visualization
2) GenomicRanges to build object read by Gviz 
3) biomaRt to retreive coordinates by gene name 
```{r setup}
library(Gviz)
library(GenomicRanges)
library(biomaRt)
library(tidyselect)
library(dplyr)
```

1.list of ts from classification RF
```{r}
classrf <- read.csv("../data/summary_analysis/Chr7_TS_master.csv",stringsAsFactors = FALSE) %>%
        dplyr::select("Gene","class_rf_score") %>% filter(!is.na(class_rf_score))
classrf.ts <- as.vector(classrf$Gene)

#change the range of frequency column to 0-1
classrf$class_rf_score = (classrf$class_rf_score - 100)/100
classrf = classrf %>% mutate(ts_predicted = class_rf_score > 0.62)
```


2.List of Experimental hits, ranked by combined score
```{r}
exp_hits <- read.csv("../data/summary_analysis/Chr7_TS_master.csv",stringsAsFactors = FALSE)
str(exp_hits)
exp.hits <- exp_hits %>%
        dplyr::select("Gene","exp_comb_score") %>%
        filter(!is.na(exp_comb_score))
exp.hits <- as.vector(exp.hits$Gene)
```


Write a function to obtain overlaps between two gene sets
```{r}
overlap <- function(list1, list2){
        overlap <- list1[(list1 %in% list2)]
        return(overlap)
}

#overlap between experimental hits and ML classification prediction
ML_class_exp_overlap <- overlap(classrf.ts,exp.hits)
ML_class_exp_overlap
```

define a function to retrieve chromosomal coordinate information from biomaRt
```{r biomaRt function}
data_retrieve <- function(gene, mart) {
  
  # input a vector of gene names
  # return a data frame containing:
  # 1) gene name: "hgnc_symbol"
  # 2) chromosome: "chromosome_name"
  # 3) start coordinate: "start_position"
  # 4) end coordinate: "end_position"
  # 5) strand: "strand"

  annotLookup <- getBM(
    mart = mart,
    attributes = c(
      "hgnc_symbol",
      "chromosome_name",
      "start_position",
      "end_position",
      "strand"
      ),
    filter = "hgnc_symbol",
    values = gene,
    uniqueRows=TRUE)
  
  #filter rows that have 1) chromosome_name == 7 and 2) no duplicates
  result <- annotLookup %>% filter(chromosome_name == 7)
  result <- result %>% filter (!duplicated(result$hgnc_symbol))
  return(result)
}
```


```{r initalize mart object}
#initialize mart object
mart.hs <- useEnsembl(biomart = "ensembl", 
                      dataset = "hsapiens_gene_ensembl")
```


```{r alternative initalization of mart using a mirror server}
#Use ensembl mirror site if emsembl not responding
mart.hs <- useEnsembl(biomart = "ensembl", 
                      dataset = "hsapiens_gene_ensembl", 
                      mirror = "uswest")
#Valid mirror options are 'www', 'uswest', 'useast', 'asia'
```


```{r Create the result dataframes from the lists}
#create result dataframe from the experiment hits 
exp.hits.result <- data_retrieve(exp.hits,mart.hs)
#attach the combined z score
exp_hits <- dplyr::rename(exp_hits,hgnc_symbol = Gene)
#map Z_comb score to range [0, 1]
exp_hits$exp_comb_score = exp_hits$exp_comb_score/max(exp_hits$exp_comb_score, na.rm = TRUE)
exp.hits.result <- exp.hits.result %>% inner_join(exp_hits[c("hgnc_symbol","exp_comb_score")],by = "hgnc_symbol")

#create the result dataframe from ML classification hits
classrf.hits.result <- data_retrieve(classrf.ts,mart.hs)
#attach the ML score 
classrf <- dplyr::rename(classrf,hgnc_symbol = Gene)
classrf.hits.result <- classrf.hits.result %>% inner_join(classrf,by = "hgnc_symbol")

#create the result dataframe with genes overlapped in the classrf and exp list and both score
overlap.result <- exp.hits.result %>% inner_join(classrf, by = "hgnc_symbol")
overlap.result = overlap.result %>% arrange(start_position)

```

```{r gene_density function}
gene_density <- function(bin_size, genedf) {
  
  # This function takes 
  # 1) bin_size in Mb
  # 2) a data.frame with gene name, start position and end position
 
  # And returns
  # 1) an array of starting positions of each bin
  # 2) an array of ending positions of each bin
  # 3) an array of gene density (# of genes/Mb) of each bin
  
  chr7_length <- 159345973
  bin_size.b <- bin_size * 1e6
  
  #calculate num of bins, n
  bin_num = ceiling(chr7_length / bin_size.b)
  
  #for Bin 1 to n-1, it is of size given, the last bin is smaller than size given
  start_array <- 0:(bin_num - 1) * bin_size.b + 1
  end_array <- c(1:(bin_num - 1) * bin_size.b, chr7_length)
  
  gene_num = rep(0, bin_num)
  genedf = genedf %>% mutate(mid_position = start_position + floor(end_position - start_position)/2, gene_bin = ceiling(mid_position/bin_size.b))
  for (i in 1:nrow(genedf)) {
    curr_row = genedf[i, ]
    gene_num[curr_row$gene_bin] = gene_num[curr_row$gene_bin] + 1
    }
  
  density = c(gene_num[1:(bin_num - 1)] / bin_size, gene_num[bin_num] * 1e6 /(end_array[bin_num] - end_array[bin_num - 1]))
  
  result = data.frame(start = start_array, end = end_array, num = gene_num, density = density)
}
```


```{r Original Graph, fig.width= 8, fig.height = 6, dpi=300}
chr <- 'chr7'
gen <- 'hg38'
#our two result dataframes are:
#exp.hits.result
#classrf.hits.result

gene_density_result <- gene_density(1, classrf.hits.result %>% select(hgnc_symbol, start_position, end_position, ts_predicted))



#create the GRange object for ML result
data_classrf <- with(classrf.hits.result, GRanges(chr, IRanges(classrf.hits.result$start_position,
                                                               classrf.hits.result$end_position),
                                                  strand = "*",
                                                  genome = gen,
                                                  #classrf.hits.result$strand,
                                                  count = classrf.hits.result$class_rf_score))

#create a GRange object for experimental hits result
data_exp <- with(exp.hits.result, GRanges(chr, IRanges(exp.hits.result$start_position,
                                                       exp.hits.result$end_position),
                                                          strand = "*",
                                                          genome = gen,
                                                          #exp.hits.result$strand,
                                                          count = exp.hits.result$exp_comb_score))

data_overlap <- with(overlap.result, GRanges(chr, IRanges(overlap.result$start_position,
                                                          overlap.result$end_position),
                                             strand = "*",
                                             genome = gen,
                                             class_rf_score = overlap.result$class_rf_score,
                                             exp_comb_score = overlap.result$exp_comb_score))

atrack1 <- AnnotationTrack(data_classrf,name = "Genes")
#atrack2 <-AnnotationTrack(data_exp,name = "Genes")

gtrack <- GenomeAxisTrack()

itrack <- IdeogramTrack(genome = gen, chromosome = chr)
                        
dtrack1 <- DataTrack(data_classrf, name = "Classification",
#                     groups = factor("Classification", 
#                                     levels = c("Classification", "Experimental Combined")),
#                     legend = TRUE, 
                     type = "p")
dtrack2 <- DataTrack(data_exp, name = "Exp. Combined",
#                     groups = factor("Experimental Combined", 
#                                     levels = c("Classification", "Experimental Combined")),
#                     legend = TRUE, 
                     type = "p", col = "magenta")
dtrack3 <- DataTrack(data_overlap, name = ("Overlayed"),
                     groups = rep(c("class_rf_score", "exp_comb_score")),
                     type = c("a", "p"))



dTrack.density <- DataTrack(start = gene_density_result$start,
                        end = gene_density_result$end,
                        chromosome = "chr7", 
                        genome = "hg38", 
                        name = "Gene Density",
                        data = gene_density_result$density,
                        type="gradient")

#ot <- OverlayTrack(trackList=list(dtrack1, dtrack2))


cdr_region <-data.frame(start = c(100117325, 138369119, 144345422),
                          end = c(102290000, 140078203, 149287349))

ht1 <- HighlightTrack(trackList=list(#atrack1, 
  dtrack1, dtrack2, dtrack3), 
                      start = cdr_region$start, 
                      end = cdr_region$end,
                      chromosome = chr,
                      genome = gen)

jpeg("../data/gviz_visualization/gviz_graph.jpeg", 
     width=12, height=9,res = 300,units = "in")
plotTracks(list(itrack, gtrack, dTrack.density, ht1))
dev.off()

tiff("../data/gviz_visualization/gviz_graph.tiff", 
     width=12, height=9,res = 300,units = "in")
plotTracks(list(itrack, gtrack, dTrack.density, ht1))
dev.off()
```

```{r}
save_pheatmap_jpeg <- function(x, filename, width=7, height=7,res = 300) {
   stopifnot(!missing(x))
   stopifnot(!missing(filename))
   jpeg(filename, width=width, height=height,res = res,units = "in")
   grid::grid.newpage()
   grid::grid.draw(x$gtable)
   dev.off()
}g
```

```{r assign CDR to each gene}
check_cdr <- function(start_position, end_position, cdr_region, cdr_num) {
  cdr_start = cdr_region[cdr_num, 'start']
  cdr_end = cdr_region[cdr_num, 'end']
  if (start_position >= cdr_start & end_position <= cdr_end) {
    return (TRUE)
  } else if ((start_position <= cdr_start & end_position >= cdr_start)) {
    return (TRUE)
  } else if ((start_position <= cdr_end & end_position >= cdr_end)) {
    return (TRUE)
  } else if ((start_position <= cdr_start & end_position >= cdr_end)) {
    return (TRUE)
  } else {
    return (FALSE)
      }
}

assign_cdr <- function(df) {
  cdr_region <-data.frame(start = c(100117325, 138369119, 144345422),
                          end = c(102290000, 140078203, 149287349))
  cdr = rep(0, nrow(df))
  
   for (i in 1:nrow(df)) {
     curr_row = df[i, ]
     curr_start = curr_row$start_position
     curr_end = curr_row$end_position
     if (curr_end < cdr_region[1, 'start'] | curr_start > cdr_region[3, 'end']) {
       cdr[i] = 0
       #print(0)
     } else if (check_cdr(curr_start, curr_end, cdr_region, 1)) {
       cdr[i] = 1
     } else if (check_cdr(curr_start, curr_end, cdr_region, 2)) {
       cdr[i] = 2
       print
     } else if (check_cdr(curr_start, curr_end, cdr_region, 3)) {
       cdr[i] = 3
     } else {
       cdr[i] = 0
     }
     if (curr_row$hgnc_symbol %in% c("HEPACAM2", "SAMD9", "SAMD9L","VPS50")) {
       cdr[i] = 4
     }
   }
  df$cdr = cdr
  return(df)
}
```


```{r}
classrf.hits.result.w.cdr = assign_cdr(classrf.hits.result) %>%
  mutate(cdr_status = (cdr != 0))
```



```{r}
mw_test <- function(cdr_num, n, df) {
  pvalue = cdr = rep(0, n)
  cdr.null <- df %>% subset(cdr == 0)
  
  if (cdr_num == 4) {
    cdr.true <- df %>% subset(cdr != 0)
  } else {
    cdr.true <- df %>% subset(cdr == cdr_num)
  }
  for(i in 1:n) {
      cdr.null.sample <- sample_n(cdr.null, nrow(cdr.true))
      pvalue[i] = wilcox.test(cdr.null.sample$class_rf_score,
                              cdr.true$class_rf_score,
                              correct=FALSE,
                              alternative = "less")$p.value
  }
  
  mean_pvalue = mean(pvalue, na.rm = TRUE)
  if (cdr_num == 4) {
    print(paste("The average p_value for ", n, "randomized Mann-Whitney Wilcoxon Ranked Sum Test of Non-CDR vs. all CDR is ", mean_pvalue, ".", sep = ""))
  } else {
    print(paste("The average p_value for ", n, " randomized Mann-Whitney Wilcoxon Ranked Sum Test of Non-CDR vs. CDR", cdr_num, " is ", mean_pvalue, ".", sep = ""))
  }
}
```


```{r}
mw_test(1, 500, classrf.hits.result.w.cdr)
```

```{r, warning= FALSE}
mw_test(2, 500, classrf.hits.result.w.cdr)
```

```{r, warning= FALSE}
mw_test(3, 500, classrf.hits.result.w.cdr)
```

```{r}
mw_test(4, 500, classrf.hits.result.w.cdr)
```


```{r}
kw_test <- function(cdr_num, n, df) {
  pvalue = cdr = rep(0, n)
  cdr.null <- df %>% subset(cdr == 0)
  if (cdr_num == 4) {
    cdr.true <- df %>% subset(cdr != 0)
  } else {
    cdr.true <- df %>% subset(cdr == cdr_num)
  }
  for(i in 1:n) {
      cdr.null.sample <- sample_n(cdr.null, nrow(cdr.true))
      test_df = rbind(cdr.null, cdr.true)
      pvalue[i] = kruskal.test(class_rf_score ~ cdr_status,
                               data = test_df)$p.value
  }
  
  mean_pvalue = mean(pvalue, na.rm = TRUE)
  if (cdr_num == 4) {
    print(paste("The average p_value for ", n, " randomized Kruskal-Wallis Ranked Sum Test of Non-CDR vs. all CDR is ", mean_pvalue, ".", sep = ""))
  } else {
    print(paste("The average p_value for ", n, " randomized Kruskal-Wallis Ranked Sum Test of Non-CDR vs. CDR", cdr_num, " is ", mean_pvalue, ".", sep = ""))
  }
  return(mean_pvalue)
}
```


```{r}
set.seed(500)
kw_test(1, 500, classrf.hits.result.w.cdr)
```

```{r}
set.seed(500)
kw_test(2, 500, classrf.hits.result.w.cdr)
```

```{r}
set.seed(500)
kw_test(3, 500, classrf.hits.result.w.cdr)
```


```{r}
set.seed(500)
kw_test(4, 500, classrf.hits.result.w.cdr)
```


```{r}
#cdr1 vs non cdr
my_test <- classrf.hits.result.w.cdr %>% subset(cdr == 0 | cdr == 1)
chisq.test(my_test$cdr_status, 
           my_test$class_rf_score)
```

```{r}
#cdr2 vs non cdr
my_test <- classrf.hits.result.w.cdr %>% subset(cdr == 0 | cdr == 2)
chisq.test(my_test$cdr_status, 
           my_test$class_rf_score)
```

```{r}
#cdr3 vs non cdr
my_test <- classrf.hits.result.w.cdr %>% subset(cdr == 0 | cdr == 3)
chisq.test(my_test$cdr_status, 
           my_test$class_rf_score)
```

```{r}
#all cdr vs non cdr
my_test <- classrf.hits.result.w.cdr
chisq.test(my_test$cdr_status, 
           my_test$class_rf_score)
```


```{r score normalization function}
score_normalization <- function(df, colname, gene_num, bin_size) {
  
  chr7_length <- 159345973
  bin_size.b <- bin_size * 1e6
  bin_num = ceiling(chr7_length / bin_size.b)
  
  df_temp = df %>% mutate(mid_position = start_position + floor(end_position - start_position)/2, bin = ceiling(mid_position/bin_size.b))
  
  normalized_score = rep(0, nrow(df))
  
  for (i in 1:nrow(df)) {
    curr_row = df_temp[i, ]
    curr_norm_score = curr_row[, colname] / gene_num[curr_row$bin]
    # if (is.na(curr_norm_score)) {
    #   print(i)
    #   print(df_temp[i,colname])
    #   print(gene_num[df_temp[i, "bin"]])
    # }
    normalized_score[i] = curr_norm_score
  }
  
  new_name = paste(colname,"_norm", sep = "")
  df_new = df %>% mutate(norm = normalized_score)
  names(df_new)[names(df_new) == 'norm'] <- new_name
  return (df_new)
}
```


```{r Normalized Graph, fig.width= 18, fig.height = 15}
chr <- 'chr7'
gen <- 'hg38'
#our two result dataframes are:
#exp.hits.result
#classrf.hits.result

gene_density_result <- gene_density(1, classrf.hits.result %>% select(hgnc_symbol, start_position, end_position))

#Normalization
classrf.hits.result.normalized = score_normalization(classrf.hits.result, "class_rf_score", gene_density_result$num, 1)
exp.hits.result.normalized = score_normalization(exp.hits.result, "exp_comb_score", gene_density_result$num, 1)
overlap.result.normalized =  score_normalization(overlap.result, "class_rf_score", gene_density_result$num, 1)
overlap.result.normalized =  score_normalization(overlap.result.normalized, "exp_comb_score", gene_density_result$num, 1)

#create the GRange object for ML result
data_classrf_norm <- with(classrf.hits.result.normalized, GRanges(chr, IRanges(classrf.hits.result.normalized$start_position,
                                                                               classrf.hits.result.normalized$end_position),
                                                                  strand = "*",
                                                                  genome = gen,
                                                                  #classrf.hits.result$strand,
                                                                  count = classrf.hits.result.normalized$class_rf_score_norm))

#create a GRange object for experimental hits result
data_exp_norm <- with(exp.hits.result.normalized, GRanges(chr, IRanges(exp.hits.result.normalized$start_position,
                                                                       exp.hits.result.normalized$end_position),
                                                          strand = "*",
                                                          genome = gen,
                                                          #exp.hits.result$strand,
                                                          count = exp.hits.result.normalized$exp_comb_score_norm))

data_overlap_norm <- with(overlap.result.normalized, GRanges(chr, IRanges(overlap.result.normalized$start_position,
                                                                          overlap.result.normalized$end_position),
                                                             strand = "*",
                                                             genome = gen,
                                                             class_rf_score = overlap.result.normalized$class_rf_score_norm,
                                                             exp_comb_score = overlap.result.normalized$exp_comb_score_norm))

atrack1 <- AnnotationTrack(data_classrf,name = "Genes")

gtrack <- GenomeAxisTrack()

itrack <- IdeogramTrack(genome = gen, chromosome = chr)
                        
dtrack1 <- DataTrack(data_classrf_norm, name = "Classification",
#                     groups = factor("Classification", 
#                                     levels = c("Classification", "Experimental Combined")),
#                     legend = TRUE, 
                     type = "p")
dtrack2 <- DataTrack(data_exp_norm, name = "Exp. Combined",
#                     groups = factor("Experimental Combined", 
#                                     levels = c("Classification", "Experimental Combined")),
#                     legend = TRUE, 
                     type = "p", col = "magenta")
dtrack3 <- DataTrack(data_overlap_norm, name = ("Overlayed"),
                     groups = rep(c("class_rf_score", "exp_comb_score")),
                     type = c("a", "p"))



dTrack.density <- DataTrack(start = gene_density_result$start,
                        end = gene_density_result$end,
                        chromosome = "chr7", 
                        genome = "hg38", 
                        name = "Gene Density",
                        data = gene_density_result$density,
                        type="gradient")
#ot <- OverlayTrack(trackList=list(dtrack1, dtrack2))


cdr_region <-data.frame(start = c(138369119, 100117325, 144345422),
                        end = c(140078203, 102627325, 148972761))

ht1 <- HighlightTrack(trackList=list(#atrack1, 
  dtrack1, dtrack2, dtrack3), 
                      start = cdr_region$start, 
                      end = cdr_region$end,
                      chromosome = chr,
                      genome = gen)

plotTracks(list(itrack, gtrack, dTrack.density, ht1))
```

