---
title: "combined_z_score"
author: "Weihan Liu"
date: "10/05/2020"
output: pdf_document
        toc = TRUE
---
1.Combine the by-plate proliferation and Erythrocyte differentiation z scores into a unified Z score
2.

```{r,results='hide'}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readxl)
library(readr)
library(tibble)
```


#combined the by-plate proliferation and Erythrocyte differentiation z scores into a unified Z score

##Read in files comtaining raw count of proliferation after normalizing to the interpolated ladder, and the erythrocyte differention raw data(this is a Ery marker ratio, so no interpolation normalization)
```{r}
pro.byplate <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/z_value_calc/edited_pro_raw_interpolated_normalized.csv",stringsAsFactors = FALSE)
ery.byplate <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/z_value_calc/edited_ediff_raw.csv",stringsAsFactors = FALSE)
```

##do a PCA analysis to see if there's batch effect between plates
```{r}
library(ggfortify)
#replace every NA values with 0, just for PCA plotting 
ery.byplate.pca.df <- ery.byplate
ery.byplate.pca.df[is.na(ery.byplate.pca.df)] = 0
pro.byplate.pca.df<- pro.byplate
pro.byplate.pca.df[is.na(pro.byplate.pca.df)] = 0

pca_ery <- prcomp(ery.byplate.pca.df[2:9], scale. =  TRUE)
pca_pro <- prcomp(pro.byplate.pca.df[2:(ncol(pro.byplate.pca.df)-1)],scale. = TRUE)
autoplot(pca_ery)
autoplot(pca_pro)
```


## For erythrocyte differentiation score, Within each plate, normalize the raw data for each gene to AAVS1 by substracting each gene's readout value - AAVS1's value. 
In this case, AAVS1 will have a value of 0
```{r}
for (i in 2:ncol(ery.byplate)){
        ery.byplate[i] = ery.byplate[i] - ery.byplate[1,i]
}
ery.byplate <- ery.byplate %>% mutate(avg_exp <- rowMeans(ery.byplate[2:9],na.rm = TRUE))
#revert the sign of the average erythrocyte score for combining later. The negative ery scores here means the erythrocyte differentiation is down after KO the gene, so these are putative myeloid TS. We want to revert their score to positive, just to be easier to combine with the proliferation score later.
ery.byplate$avg_exp <- -ery.byplate$avg_exp
```


## For proliferation scores, within each plate, normalize the interpolated data to AAVS1
```{r}
for (i in 2:ncol(pro.byplate)){
        pro.byplate[i] = pro.byplate[i] - pro.byplate[1,i]
}
pro.byplate <- pro.byplate %>% mutate(avg_pro <- rowMeans(pro.byplate[2:ncol(pro.byplate)],na.rm = TRUE))
pro.byplate <- rename(pro.byplate,avg.pro = "avg_pro <- rowMeans(pro.byplate[2:ncol(pro.byplate)], na.rm = TRUE)")
```


combine the proliferation and erythrocyte average score with the z scores into a new dataframe
```{r}
comb_df <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/z_value_calc/edited_both_z_avg_exp_comb.csv",stringsAsFactors = FALSE)
comb_df <- comb_df %>% inner_join(pro.byplate[,c("Name","avg.pro")],by = "Name") %>%
        inner_join(ery.byplate[,c("Name","avg_exp")],by = "Name")


rownames(comb_df) <- comb_df[,1]
comb_df <- select(comb_df,-"Name")
#normalize your data onto 0:1 range, to get rid of directionality
normalize <- function(df){
        apply(df,2,function(x) (x - min(x))/(max(x)-min(x)))
}
        
comb_df_norm <- as.data.frame(normalize(comb_df))

#create combined z score and avg score columns
comb_df_norm$z_comb <- comb_df_norm$Pro.z+comb_df_norm$E.diff.z
#comb_df_norm$avg_comb <- comb_df_norm$
comb_df_norm$avg_comb <- comb_df_norm$avg.pro + comb_df_norm$avg.ery

View(comb_df)
View(comb_df_norm)

str(comb_df_norm)
```

save the file
```{r}
write.csv(comb_df_norm,"/Users/weihan/Documents/GitHub/ts_machine_learning/data/z_value_calc/comb_ranking.csv")
```

We decided to use the average normalized expression instead of the z score method to score the genes. So now let's calculate the significance value of the experimental genes vs AAVS1 control

1.In experimental design, each gene KO is performed in a separate well, so we can treat genes as independent. We can perform statistical test of each gene vs AAVS1 respectively.

2.We need to perform stat test for proliferation and erythrocyte differentiation respectively, since they are different biological entities.

3.Sample size is small(#of replicate), so we cannot assume their distribution is normal. Therefore we will use non-parametric Mann-Whitney U test

4.We will perform the test on the normalized(gene - AAVS1) and min-max transformed data. This data is what we used for the final ranking
```{r}
#create a copy data for significance test
ery.byplate.test <- ery.byplate %>%
        select(-c("avg_exp <- rowMeans(ery.byplate[2:9], na.rm = TRUE)","avg_exp"))
pro.byplate.test <- select(pro.byplate,-"avg.pro")
#revert the sign of all erththrocyte differentiation reads here. We are interested in the negative erythrocyte differentiation, since they mean after KO the ery differentiation is down, they are the putative TS genes. Let's make them positive
#vector of AAVS1 erythrocyte 
ery.byplate.test[,2:ncol(ery.byplate.test)] <- -ery.byplate.test[,2:ncol(ery.byplate.test)]
```

Write a function to perform min max normalization
```{r}
normalize <- function(df){
        apply(df,2,function(x) (x - min(na.omit(x)))/(max(na.omit(x))-min(na.omit(x))))
}
```

```{r}
ery.byplate.test.norm <- as.data.frame(normalize(ery.byplate.test))
pro.byplate.test.norm <- as.data.frame(normalize(pro.byplate.test))

ery.byplate.test.norm <- rownames_to_column(ery.byplate.test.norm,var = "Gene")
pro.byplate.test.norm <- rownames_to_column(pro.byplate.test.norm ,var = "Gene")
```


Significance test to determine the genes significantly different from AAVS1


parametric t test
```{r}
#because t test assume sample distribution, we have to perform the t test on proliferation and erythrocyte differentiation seperately

#define a function to perform student t test for each gene comparing to AAVS1 and store the resulting p value in a new column
t_test <- function(df,alpha){ 
        rownames_df <- rownames(df)
        df <- mutate(df,t_test_p = NA)
        AAVS1 <- as.numeric(df[1,])
        for (i in 1:nrow(df)) {
            x <- t.test(AAVS1,as.numeric(df[i,]),alternative = "less")
            df[i,"t_test_p"] <- x$p.value
        }
        row.names(df) <- rownames_df
        hits <- filter(df,t_test_p < alpha)
        return(list(df,hits))
       
```


```{r}
ery.byplate.test.p <- t_test(ery.byplate.test.norm,0.1)[[1]]
ery.hits <- t_test(ery.byplate.test.norm,0.1)[[2]]

pro.byplate.test.p <- t_test(pro.byplate.test.norm,0.01)[[1]]
pro.hits <- t_test(pro.byplate.test.norm,0.01)[[2]]
```

#non-parametric Mann-Whitney test 
```{r}
#merge the proliferation and erythrocyte table
pro.ery.merged <- full_join(ery.byplate.test.norm,pro.byplate.test.norm,by = "Gene")
pro.ery.merged <- column_to_rownames(pro.ery.merged,var = "Gene")

mw_test <- function(df,alpha){ 
        rownames_df <- rownames(df)
        df <- mutate(df,t_test_p = NA)
        AAVS1 <- as.numeric(df[1,])
        for (i in 1:nrow(df)) {
            x <- wilcox.test(AAVS1,as.numeric(df[i,]),alternative = "less")
            df[i,"t_test_p"] <- x$p.value
        }
        row.names(df) <- rownames_df
        hits <- filter(df,t_test_p < alpha)
        return(list(df,hits))
}

pro.ery.merged.p <- mw_test(pro.ery.merged,0.01)[[1]]
#the significant genes when performing proliferation and erythrocyte score together        
pro.ery.hits <- rownames(mw_test(pro.ery.merged,0.01)[[2]]) 
```

 

Plotting
```{r}
comb_df_norm <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/z_value_calc/comb_ranking.csv",stringsAsFactors = FALSE)

library(ggrepel)
#obtain the top ranked genes by both z score and 
comb_df_norm <- rownames_to_column(comb_df_norm, var = "Gene")


# #label the top 20% gene by combined z score
# highlight_z_comb_df <- top_frac(comb_df_norm,0.20,z_comb)
# comb_df_norm %>% ggplot(aes(x = Pro.z,y = E.diff.z)) + 
#         geom_point(alpha = 0.3) +
#         geom_point(data=highlight_z_comb_df, aes(x = Pro.z,y = E.diff.z), color='red',size=2)+
#         geom_text_repel(aes(label=ifelse(comb_df_norm$z_comb %in% (top_frac(comb_df_norm,0.20,z_comb)$z_comb),as.character(Gene),''))) +
#         ggtitle("Top 20% genes ranked by combined z scores") +
#         labs(x = "proliferation z score", y = "erythrocyte differentiation z score") +
#         theme_bw() +
#         theme(plot.title = element_text(hjust = 0.5),panel.border = element_blank(), panel.grid.major = element_blank(),
#         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) 
#        

#label the top 20% gene by combined average expression
highlight_avg_exp_comb_df <- top_frac(comb_df_norm,0.2,avg_comb)

#label the significant genes detected by Mann-Whitney test on combined proliferation and erythrocyte differentiation score
sig_genes_highlight <- filter(comb_df_norm,Gene %in% pro.ery.hits)
#select out AAVS1 for highlighting
AAVS1_highlight <- filter(comb_df_norm,Gene == "AAVS1")

comb_df_norm %>% ggplot(aes(x = avg.pro,y = avg.ery)) + 
        geom_point(alpha = 0.3) +
        geom_point(data = sig_genes_highlight, aes(x = avg.pro,y = avg.ery), color = "red", size = 2)+
        geom_point(data = AAVS1_highlight, aes(x = avg.pro,y = avg.ery), color = "blue", size = 2)+
        geom_text_repel(aes(label=ifelse(comb_df_norm$avg_comb %in% (top_frac(comb_df_norm,0.20,avg_comb)$avg_comb) | comb_df_norm$Gene == "AAVS1",as.character(Gene),''))) +
        #ggtitle("Top 20% genes ranked by AAVS1 normalized combined values") +
        labs(x = "average proliferation", y = "average erythrocyte differentiation") +
        theme_bw() +
        theme(plot.title = element_text(hjust = 0.5),panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) 
        

```


Statistical Test on the significant genes

#.Statistical test on whether jeremy's experimental gene sets bias towards proliferation
I chose the Mann-Whitney U test

##Here we are comparing two sets of data:
1.proliferation reading of AAVS1
2.proliferation reading of the average of all other genes
```{r}
#read in the proliferation data
prol <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/z_value_calc/edited_pro_raw_interpolated_normalized.csv",stringsAsFactors = FALSE)

#proliferation reads of AAVS1 in each plate
AAVS1 <- as.numeric(prol[1,2:ncol(prol)])

#average proliferation reads of all other genes in each plate 
all_genes <- as.numeric(colMeans(prol[2:nrow(prol),2:ncol(prol)],na.rm = TRUE))

#Mann-Witney test comparing avg proliferation of AAVS1 vs all other experimental genes
wilcox.test(AAVS1,all_genes,alternative = "less")

```


##compare
1.proliferation reading of AAVS1
2.proliferation reading of the average of other experimental genes that are on CDR
```{r}
#average proliferation reads of all CDR genes in Jeremy's experiment
CDR <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/CDR_list.csv",stringsAsFactors = FALSE)
all_genes_cdr <- prol %>% 
        filter(Name %in% CDR$Name)
all_genes_cdr <- colMeans(all_genes_cdr[,2:ncol(all_genes_cdr)],na.rm = TRUE) %>% 
        as.numeric()

#Mann-Witney test comparing avg proliferation of AAVS1 vs all other experimental genes that are on CDR
wilcox.test(AAVS1,all_genes_cdr,alternative = "less")
```


##compare
1.proliferation reading of AAVS1
2.proliferation reading of the average of other experimental genes that are ML classification hits
```{r}
#read in the predicted TS hits from classification RF
class_rf_hits <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/gwide_hema_classification/pred_hits_CGC_bootstrap.csv",stringsAsFactors = FALSE)

#subset the top 120 genes as top hits and let's see which experimental genes overlap with these top hits
all_genes_ml_overlap <- prol %>% 
        filter(Name %in% class_rf_hits[1:110,]$Gene)

#calculate the plate averages of these experimental genes overlapped with ML hits
all_genes_ml_overlap <- colMeans(all_genes_ml_overlap[,2:ncol(all_genes_ml_overlap)],na.rm = TRUE) %>%
        as.numeric()

#perform Mann-Whitney test
wilcox.test(AAVS1,all_genes_ml_overlap,alternative = "less")
```



