---
title: "gwide_random_forest_implementation"
author: "Weihan Liu"
date: "17/02/2020"
output: html_document
---

Pseudo code for regression random forest implementation:

1.  Given training data set
2.  Select number of trees to build (ntrees)
3.  for i = 1 to ntrees do
4.  |  Generate a bootstrap sample of the original data
5.  |  Grow a regression tree to the bootstrapped data
6.  |  for each split do
7.  |  | Select m variables at random from all p variables
8.  |  | Pick the best variable/split-point among the m
9.  |  | Split the node into two child nodes
10. |  end
11. | Use typical tree model stopping criteria to determine when a tree is complete (but do not prune)
12. end

Install relevant packages
```{r}

```


Read in files
```{r}
gwide <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/gwide_hema_classification/gwide_dup_rm.csv")  #duplicated removed
gwide_knn <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/gwide_hema_classification/gwide_dup_rm_knn.csv) #duplicated removed + knn imputated
```

