---
title: "Re-run with AML data only"
author: "Weihan Liu"
date: "15/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Table of contents:
1.data cleaning: Filter columns based on NA and 0 values and perform KNN imputation
2.Create train, test and validation set

```{r}
library(tidyselect)
library(data.table)
library(dplyr)
```


#1 data cleaning
Load in data. This table contains screen results from hematological cancers, and only protein coding genes
```{r}
gwide <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/gwide_hema_classification/gwide.csv", stringsAsFactors = FALSE)

#drop the column 
#1)Weissman2014.CRISPRi.Growth.phenotype..mean.of.top.3.gammas. because most of the values are underpowered to detect tumor suppressors in CRISPRi screen, 
#2)HSC.logCPM column
#3)all the threshold columns, these are columns containing 0 opr 1s based on our judgement of whether they pass the thrshold

gwide <- gwide %>% 
        select(-c("HSC.logCPM","Weissman2014.CRISPRi.Growth.phenotype..mean.of.top.3.gammas.","X","X.1","X.2","GeneID"))
                    
#get rid of all the threshold columns
 
threshold_vars <- vars_select(colnames(gwide),ends_with("threshold"))
gwide <- gwide %>% select(-threshold_vars) %>% glimpse()

#remove all duplicated values in gwide.unique() function remove the rows whose all columns are identical(because I didn't specify "by =" here, so the removal of duplicated rows are based on all columns)
gwide <- unique(gwide)
unique(gwide,by = "Gene") 
#back-up original data
gwide_original <- gwide
#there are still 22 some duplicated genes by gene name(but some other columns differ, so they didn't get removed in the previous command, so let's do another round of removing duplicate by Gene name
gwide <- gwide[!duplicated(gwide$Gene),]
gwide_original <- gwide
glimpse(gwide)
```

##add TSgene ground truth column
```{r}
#clean the tsgene dataframe
tsgene <- read.csv("/Users/weihan/Documents/GitHub/ts_machine_learning/data/gwide_hema_classification/TSGene_all.csv",stringsAsFactors = FALSE)
head(tsgene)
table(tsgene$GeneType)
tsgene <- tsgene %>% filter(GeneType == "protein-coding") %>% 
        select("GeneSymbol") %>%
        rename("Gene" = "GeneSymbol") %>%
        mutate("TS_status" = 1)
#join tsgene and the gwide dataframe

gwide <- full_join(gwide,tsgene,by = "Gene")
gwide$TS_status[is.na(gwide$TS_status)] <- 0 #replace all NAs in TS_status column with 0s
```

##add Elledge 2013 data columns to enrich training data
```{r}
library(readxl)
elledge <- read_xlsx("/Users/weihan/Documents/GitHub/ts_machine_learning/data/gwide_hema_classification/elledge2013supplement.xlsx")
elledge <- elledge %>% select(c("Gene","LOF/Silent*","Splicing/Benign*","MissDamaging/Benign*"))

gwide <- full_join(gwide, elledge, by = "Gene")

gwide <- gwide %>% rename(LOF_Silent = `LOF/Silent*`,
                          Splicing_Benign = `Splicing/Benign*`,
                          MissDamaging_Benign = `MissDamaging/Benign*`)
glimpse(gwide)
```

##Drop all the p value columns
```{r}
gwide <- gwide %>% select(-c(
        "Chen2019.pos.p.value",
        "Chen2019.pos.fdr",
        "Chen2019.pos.rank",
        "Doench2018.Average.nlog.p.values.",
        "Elledge2013.TUSON_p_value_TSG",
        "Elledge2013.TUSON_q_value_TSG",
        "Elledge2019.HMEC.Combined.pvalue.drop",
        "Elledge2019.HMEC.FDR.drop",
        "Elledge2019.HPNE.Combined.pvalue.drop",
        "Elledge2019.HPNE.FDR.drop",
        "Sabbatini2015.KBM7.adjusted.p.value",
        "Sabbatini2015.K562.adjusted.p.value",
        "Sabbatini2015.Jiyoye.adjusted.p.value",
        "Sabbatini2015.Raji.adjusted.p.value",
        "Wallace2016.DEseq2.P.Value",
        "Rep1_2.Mann.Whitney.p.value",
        "Rep1_2.Mann.Whitney.p.value.1"
))


```



##Check NA and 0 status for each column and get rid of outliers
```{r}
#rank NA percentage by column
sapply(gwide, function(x) sum(is.na(x))/length(x)) %>% 
        sort(decreasing = TRUE)
#get ride of the Wallace2016.DEseq2.Log2.Fold.Change too many NA values(over 90 %)
gwide <- gwide %>% 
        select(-"Wallace2016.DEseq2.Log2.Fold.Change")

#0 values
sapply(gwide, function(x) sum(x == 0,na.rm = TRUE)/length(x)) %>%
        sort(decreasing = TRUE)
```


Retain only the AML columns
```{r}
gwide <- gwide %>% select(Gene,
                          chromosome,
                          Blomen.2015.hap1.GTS.ratio = Brummelkamp.hap1.GTS.ratio,
                          Blomen.2015.kbm7.GTS.ratio = Brummelkamp.kbm7.GTS.ratio,
                          Wang.2015.KBM7.CS = Sabbatini2015.KBM7.CS,
                          Wang.2015.K562.CS = Sabbatini2015.K562.CS, 
                          Wang.2017.EOL.1 = Sabbatini2017.EOL.1,
                          Wang.2017.HEL = Sabbatini2017.HEL,
                          Wang.2017.MOLM.13 = Sabbatini2017.MOLM.13,
                          Wang.2017.MonoMac1 = Sabbatini2017.MonoMac1,
                          Wang.2017.NB4.rep1 = Sabbatini2017.NB4.rep1,
                          Wang.2017.NB4.rep2 = Sabbatini2017.NB4.rep2,
                          Wang.2017.OCI.AML2 = Sabbatini2017.OCI.AML2,
                          Wang.2017.OCI.AML3 = Sabbatini2017.OCI.AML3,
                          Wang.2017.OCI.AML5 = Sabbatini2017.OCI.AML5,
                          Wang.2017.P31.FUJ = Sabbatini2017.P31.FUJ,
                          Wang.2017.PL.21 = Sabbatini2017.PL.21,
                          Wang.2017.SKM.1 = Sabbatini2017.SKM.1,
                          Wang.2017.TF.1 = Sabbatini2017.TF.1,
                          Wang.2017.THP.1 = Sabbatini2017.THP.1,
                          Horlbeck.2016.CRISPRa.average.phenotype.of.strongest = average.phenotype.of.strongest.3.2,
                          Horlbeck.2016.CRISPRi.average.phenotype.of.strongest.3.2 = average.phenotype.of.strongest.3.2.1,
                          TS_status)

```



##knn imputation
```{r}
library(caret)
preProcValues <- preProcess(gwide %>% 
                          dplyr::select(-c("")),
                            method = c("knnImpute"),
                            k = 20,
                            knnSummary = mean)
impute_cars_info <- predict(preProcValues, cars_info,na.action = na.pass)
```





```{r}
library(data.table)
library(impute)
library(tibble)


gene_chr <- gwide[c("Gene","chromosome")]

#remove rownames from gwide
rownames(gwide) <- c()
gwide_knn <- gwide %>% column_to_rownames(var = "Gene") %>% select(-"chromosome") 

gwide_knn <- impute.knn(as.matrix(gwide_knn[ ,!names(gwide_knn) %in% "TS_status"]),k = 10)

# perform knn imputation.
anyNA(gwide_knn)
gwide_knn <- as.data.frame(gwide_knn$data)

#attach back the gene and chromosome columns
gwide_knn <- gwide_knn %>% rownames_to_column(var = "Gene") %>%
        inner_join(gene_chr, by = "Gene") 
#attach back the TS_status column
gene_ts_status <- gwide[(c("Gene","TS_status"))]
gwide_knn <- gwide_knn %>% inner_join(gene_ts_status, by = "Gene")

#we have some NA values in the TS_status column, which are genes that don't have TS vs Non TS annotation, let's replace them with 0
gwide_knn$TS_status[is.na(gwide_knn$TS_status)] <- 0

glimpse(gwide_knn)

#save the knn_imputated file 
write.csv(gwide_knn,"/Users/weihan/Documents/GitHub/ts_machine_learning/data/gwide_hema_classification/gwide_dup_rm_knn_AML_CML_only.csv")
```