---
title: "regression_RF_implement"
author: "Weihan Liu"
date: "25/03/2020"
output: html_document
---

```{r}
library(rsample)      # data splitting 
library(randomForest) # basic implementation
library(ranger)       # a faster implementation of randomForest
library(caret)        # an aggregator package for performing many machine learning models
library(h2o)   
```


We inherit the training and testing data split from the classification data preprocessing
manipulate the training and testing data, we will use the Elledge Lasso probability as our ground truth column. THE CGC TS status will be one of our predictor column
```{r}
#clean testing data
#delete the columns from Elledge 2013, except the lasso prob column, which will be used as ground truth column.We delete all the other columns from Elledge 2013 because they are correlated with
gwide_test_CGC_reg <- select(gwide_test_CGC,-c("LOF_Silent","Splicing_Benign","MissDamaging_Benign","hits"))
 
#clean training data
gwide_train_balanced_CGC_reg <- select(gwide_train_balance_CGC,-c("LOF_Silent","Splicing_Benign","MissDamaging_Benign"))
```

Hyperparameter tuning(using ranger)
```{r}
#construct hyperparameter grid
hyper_grid_reg <- expand.grid(
  mtry <- seq(2, ncol(gwide_train_balanced_CGC_reg) * 0.8, 2),
  node_size  = seq(3, 9, by = 2),
  sampe_size = c(0.6,0.7,0.8),
  OOB_RMSE   = 0
)

# total number of combinations
nrow(hyper_grid_reg)
```

Execute hyperparameter grid search
```{r}

for(i in 1:nrow(hyper_grid_reg)) {
  
  # train model
  model <- ranger(
    formula         = Elledge2013.TSG_Probability_LASSO ~ ., 
    data            = select(gwide_train_balanced_CGC_reg,-"chromosome"), 
    num.trees       = 500,
    mtry            = hyper_grid_reg$mtry[i],
    min.node.size   = hyper_grid_reg$node_size[i],
    sample.fraction = hyper_grid_reg$sampe_size[i],
    seed            = 123
  )
  
  # add OOB error to grid
  hyper_grid_reg$OOB_RMSE[i] <- sqrt(model$prediction.error)
}

c
#check out the top hyperparameters
hyper_grid_reg %>% 
  dplyr::arrange(OOB_RMSE) %>%
  head(20)

#We can choose mtry = 8, nodesize = 9, sample size = 0.8, the resulting OOB_RMSE is 0.2770774
```


plot the distribution of the OOB_RMSE
```{r}
OOB_RMSE <- vector(mode = "numeric", length = 100)

for(i in seq_along(OOB_RMSE)) {

  optimal_ranger <- ranger(
    formula         = Elledge2013.TSG_Probability_LASSO ~ ., 
    data            = select(gwide_train_balanced_CGC_reg,-"chromosome"), 
    num.trees       = 500,
    mtry            = 8,
    min.node.size   = 9,
    sample.fraction = .8,
    importance      = 'impurity'
  )
  
  OOB_RMSE[i] <- sqrt(optimal_ranger$prediction.error)
}

hist(OOB_RMSE, breaks = 20)
```

Assess variable importance
```{r}
var_imp <- data.frame(optimal_ranger$variable.importance)
var_imp <- rownames_to_column(var_imp,var = "variable")

var_imp %>% 
  dplyr::arrange(desc(var_imp$optimal_ranger.variable.importance)) %>%
  dplyr::top_n(29) %>%
  ggplot(aes(reorder(var_imp$variable, var_imp$optimal_ranger.variable.importance), var_imp$optimal_ranger.variable.importance)) +
  geom_col() +
  coord_flip() +
  ggtitle("Top 25 important variables")


```

```{r,fig.height=4, fig.width=8}
#Random Forest
# rf_reg = randomForest(Elledge2013.TSG_Probability_LASSO ~ .,  
#                    data = select(gwide_train_balanced_CGC_reg,-"chromosome"),importance = TRUE,mtry = 8,nodesize = 9,sampsize = nrow(gwide_train_balanced_CGC_reg))
# plot(rf_reg) 
# print(rf_reg)
# varImpPlot(rf_reg)



#ranger
rf_reg_ranger <- ranger(
    formula   = Elledge2013.TSG_Probability_LASSO ~ ., 
    data      = select(gwide_train_balanced_CGC_reg,-"chromosome"), 
    num.trees = 500,
    min.node.size = 9,
    mtry      = 8,
    sample.fraction = 0.8
  )

```

Predicting
```{r}
pred_prob <- predict(rf_reg_ranger, gwide_test_CGC_reg)
pred_prob

gwide_test_CGC_reg$pred_prob <- pred_prob$predictions

ranked_result <- gwide_test_CGC_reg[
  with(gwide_test_CGC_reg, order(pred_prob,decreasing = TRUE)),
]

ranked_result <- ranked_result %>% select("Gene","TS_status","pred_prob")
ranked_result 

write.csv(ranked_result,"/Users/weihan/Documents/GitHub/ts_machine_learning/data/gwide_hema_regression/ranked_result.csv")
```


Performance measuree
```{r}
library(PRROC)

PRROC_obj <- roc.curve(scores.class0 = gwide_test_CGC_reg$pred_prob , gwide_test_CGC_reg$Elledge2013.TSG_Probability_LASSO,
                       curve=TRUE)
plot(PRROC_obj)
```

